---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stanova

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/bayesstuff/stanova.svg?branch=master)](https://travis-ci.org/bayesstuff/stanova)
<!-- badges: end -->

The goal of `stanova` is to provide a more relevant and interpretable `summary` for Bayesian models with categorical covariates and possibly interactions. At the moment, it only provides wrappers for `rstanarm` via functions `stanova_lmer` and `stanova_glmer`.

The issue `stanova` tries to address is that categorical variables with $k$ levels need to be transformed into $k-1$ numerical model coefficients. This poses a problem if a model includes a factor with more than two levels that interacts with another variable. In this case, the most reasonable parameterization of the model is such that the intercept correspond to the (unweighted) grand mean and therefore tests of main effects represent average effects (compared to simple effects). In this parameterization, factors with $k$ levels, where $k > 2$ (i.e., more than two levels), cannot be mapped in a 1-to-1 fashion to the $k-1$ coefficients as no such mapping exists. Thus, the  values of the model coefficients do not represent effects of one factor level, but always pertain to more than one factor level and thus cannot be directly interpreted. In other words, these coefficients should not be looked at. Instead, `stanova` transforms these parameters back such that one gets the information on the factor levels (or combination of factor levels for interactions). The default output shows for each factor level the difference from the intercept which as discussed before corresponds to the (unweighted) grand mean.

Another problem adressed by `stanova` is that for Bayesian models the mapping of factor-levels to model coefficients needs to be done such that the marginal prior for each factor-level is the same. If one uses one of the contrast coding schemes commonly used in frequentist models in which the intercept corresponds to the (unweighted) grand mean the marginal prior differs across factor levels. For example, when using `contr.sum` all but the last factor level are mapped to exactly one model coefficient with positive sign, and the last factor level is mapped with negative sign on all model coefficients. Thus, the marginal prior for the last factor level is more diffuse than for the other factor levels (if $k >2$). `stanova` per default uses the prior suggested by Rouder, Morey, Speckman, and Province (2012) in which the marginal prior is the same for all factor levels. In this prior, the sum-to-zero constraint that is necessary for the intercept to represent the (unweighted) grand mean is also imposed. This contrast is implemented in the `contr.bayes()` function.

## Installation

For the moment, you can only install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bayesstuff/stanova")
```
## Example

The most basic example only uses a single factor with three levels to demonstrate the output. 

```{r fit1, results='hide', message=FALSE, warning=FALSE}
library(stanova)
data("Machines", package = "MEMSS")

m_machines <- stanova_lmer(score ~ Machine + (Machine|Worker),
                           data=Machines, chains = 2, iter = 500)

```

The `summary` method of a `staonva` objects first provides some general information about the fitted model (similar to `rstanarm`). It then provides statistics about the intercept. The next block provides estimates for each factor-level of the `Machine` factor. These estimates represent the difference of the factor level against the intercept. For example, for `Machine A`, the estimate is around `-7` suggesting that the mean of this factor level is 7 points below the intercept (i.e., the grand mean). 

```{r}
summary(m_machines)
```

The key to this output is the `stanova_samples()` function which takes a fitted model objects and returns the posterior samples transformed to represent the difference from the intercept for each factor level. The default output is an `array`, but this can be changed to a `matrix` or `data.frame` with the `return` argument.

```{r}
out_array <- stanova_samples(m_machines)
str(out_array)
```

One can also change which dimension of the `array` represents the chain via the `dimension_chain` argument.
```{r}
out_array2 <- stanova_samples(m_machines, dimension_chain = 2)
str(out_array2)
```

This makes it easy to produce plots via the `bayesplot` package on the level of the differences from the intercept:

```{r, fig.width=8, fig.height=3}
bayesplot::mcmc_trace(out_array2$Machine)
```


# References

Rouder, J. N., Morey, R. D., Speckman, P. L., & Province, J. M. (2012). Default Bayes factors for ANOVA designs. Journal of Mathematical Psychology, 56(5), 356–374. https://doi.org/10.1016/j.jmp.2012.08.001

